
\documentclass{article}
%% Needs to be here 
\usepackage{hyperref}
\usepackage{amssymb}
\usepackage{amsmath}

\newcommand{\gep}{\texttt{getelementptr}}
\newcommand{\naturals}{\mathbb{N}}
\newcommand{\integers}{\mathbb{Z}}

\title{Array representations for Polly}
\author{(Anonymous)}

\usepackage[utf8]{inputenc}
\begin{document}
\maketitle

\newpage
\tableofcontents
\newpage

\section{Introduction}
We mathematically describe the array representations that are used by 
Fortran and Chapel, two programming languages which support complex array
indexing. We motivate the need for LLVM to be able to represent these
array representations and their cooresponding indexing operation at a
semantic level that is higher than that of \gep.
We showcase existing implementations of such a semantic representation,
and would like to propose this as an LLVM standard. Choices for
implementation are outlined in this document.

\section{Formalization}
We first formalize the most general type of array we have encountered - That
of Chapel style arrays which are multidimensional, strided, and provide
non-zero-based indexing.

We denote the set of all possible arrays (a loosely defined term, to be sure)
as $Array$.
\begin{align*}
    &dim: Array \to \naturals \\
    &dim(A) = \text{dimensionality of array $A$}
    \\
    \\
    &ixbegin/ixend: Array \times \lbrack 0, dim(A) \rbrack \to \integers \\
    &ixbegin/ixend(A, d) = \text{first and last legal indeces allowed along dimension $d$}
    \\
    \\
    &stride: Array \times \lbrack 0, dim(A) \rbrack \to \naturals \\
    &stride(A, d) = \text{stride of an array $A$ along dimension $d$} \\
\end{align*}

Using the data outlined above, we can now define what a correct indexing function
into an array is. An indexing function recieves \textbf{logical coordinates}
of the array index, which is converted into \textbf{physical coordinates},
which is the in-memory representation.

We now try to define a partial function $logicalToPhysical$, which maps
logical coordinates to physical coordinates.

First, we define $ixset(A, d)$, the set of valid indeces along a dimension $d$:
\begin{align*}
    &ixset: Array \times \lbrack 0, dim(A) \rbrack \to 2^\naturals\\
    &ixset(A, d) = \{ ixbegin(A, d) + \alpha \cdot stride(A, d)~\vert~ \alpha \in \naturals \} \cap (-\infty, ixend(A, d) \rbrack
\end{align*}

The $ixset$ constructs the set of valid indeces that are allowed along
a given dimension. We first give the type of $logicalToPhysical$, and
we then gradually build up the full expression.

\begin{align*}
    logicalToPhysical: Array \times \prod_{d=1}^{dim(A)} ixset(A, d) \to \naturals \\
\end{align*}

Notice that the logical coordinates live in $\prod_{d=1}^{dim(A)} ixset(A, d)$,
while physical coordinates live in $\naturals$. We wish to view memory
as a 1-D sequence of values, with $0$ being the "base address" of the array.
The logical coordinates allow us to index the "virtual view" of the array
that we posess.

\section{C-style arrays --- no stride, no offset}
In the case of C-style arrays, our $ixset(A, i) = \big\lbrack 0, N(A, i) \big) \subset \naturals$,
where $N(A, i)$ is informally the "size" of array $A$ along dimension $i$.
The word "size" is, however, fraught with danger, since it conflates
the size of the "logical indexing space" and the underlying "memory space".
In the case of the C language, both of these concepts coincide, but this
is not so in other languages such as Chapel.

In the case of a C array, here are the different functions that we have defined:
\begin{itemize}
    \renewcommand\labelitemi{--}
    \item Declaration: \texttt{T A[$N_1$][$N_2$]$\dots$[$N_d$]}
    \item Dimension: $dim(A) = d$
    \item Beginning, ending indeces: $ixbegin(A, \_) = 0$, $ixend(A, d) = N_d - 1$
    \item Stride: $stride(A, \_) = 1$
    \item Index set: 
        \begin{align*}
            ixset(A, d) &= \{ ixbegin(A, d) + \alpha \cdot stride(A, d)~\vert~ \alpha \in \naturals \} \cap (-\infty, ixend(A, d) ] \\
                        &= \{ 0 + \alpha \cdot 1~\vert~\alpha \in \naturals \} \cap (-\infty, N_d - 1 \rbrack \\
                        &= [0, \infty) \cap (-\infty, N_d - 1]\\
                        &= \lbrack 0, N_d - 1 \rbrack
        \end{align*}
\end{itemize}

To calculate the index expression, we create a function, $\#elem(A, d) = |ixset(A, d)|$. 
$\#elem$ measures the number of elements present along a given dimension.
With this ingredient, we can finally re-create the C array indexing function as:

\paragraph{Sid question: How do you actually \textbf{motivate} this definition?}

\begin{align*}
    &logicalToPhysical: Array \times \prod_{d=1}^{dim(A)} ixset(A, d) \to \naturals \\
    &logicalToPhysical(A, (ix_1, ix_2, \dots, ix_{dim(A)})) = \\
    &\sum_{d=1}^{dim(A)} ix_d \times \prod_{prevd=1}^{d - 1} \#elem(A, prevd)
\end{align*}

With respect to this formula, we understand that $ix_1$ is the innermost dimension
with an increment of 1, and $ix_{dim(A)}$ is the outermost dimension.
\section{Fortran-style arrays: only offset}
Let us next consider languages which allow us to create arrays with arbitrary
starting and ending indeces, such as Fortran.

\begin{itemize}
    \renewcommand\labelitemi{--}
    \item Declaration: \texttt{T A[$B_1:E_1$][$B_2:E_2$]$\dots$[$B_d:E_d$]}
    \item Dimension: $dim(A) = d$
    \item Beginning, ending indeces: $ixbegin(A, d) = B_d$, $ixend(A, d) = E_d$
    \item Stride: $stride(A, \_) = 1$
    \item Index set: 
        \begin{align*}
            ixset(A, d) &= \{ ixbegin(A, d) + \alpha \cdot stride(A, d)~\vert~ \alpha \in \naturals \} \cap (-\infty, ixend(A, d) ] \\
                        &= \{ B_d + \alpha \cdot 1~\vert~\alpha \in \naturals \} \cap (-\infty, E_d ] \\
                        &= [B_d, \infty) \cap (-\infty, E_d]\\
                        &= [ B_d, E_d]
        \end{align*}
\end{itemize}

In this case, the value of  $\#elem(A, d) = E_d - B_d + 1$.


\begin{align*}
    &logicalToPhysical: Array \times \prod_{d=1}^{dim(A)} ixset(A, d) \to \naturals \\
    &logicalToPhysical(A, (ix_1, ix_2, \dots, ix_{dim(A)})) = \\
    &\sum_{d=1}^{dim(A)} (ix_d - B_d) \times \prod_{prevd=1}^{d - 1} \#elem(A, prevd)
\end{align*}


\section{Chapel-style arrays: strides and offsets}

\begin{align*}
    &logicalToPhysical: Array \times \prod_{d=1}^{dim(A)} ixset(A, d) \to \naturals \\
    &logicalToPhysical(A, (ix_1, ix_2, \dots, ix_{dim(A)})) = \\
    &\sum_{d=1}^{dim(A)} ((ix_d - B_d) / stride(A, i)) \times \prod_{prevd=1}^{d - 1} \#elem(A, prevd)
\end{align*}

\section{Reduction of representation}
It seems like we require a lot of information for even a single array:
that of $dim$, $ixbegin$, $ixend$ and $stride$.  We first show
to eliminate the requirement of a per-dimension $ixbegin$ and $ixend$. We
change this information into a per-dimension $\#elem$, and a 
\textbf{single $offset$ value for an entire array}. 

\section{Multidimensional representation for LLVM}
We have encoded the final representation of arrays in two experimental projects -
one being a fork of LLVM + dragonegg + Polly, and the other a fork of
LLVM + Chapel + Polly. The first implements this style of indexing in Fortran,
which is interpreted by Polly for polyhedral optimisation, and the second
project does the same for Chapel.

We argue that this representation will provide more accurate analysis and
optimisation opportunities for LLVM, due to the inherent loss in information
by the linearization performed by \gep. For a detailed reference of 
the loss of information and the (\textit{necessarily incomplete}) heuristics
that are used to regain this information, refer to \cite{grosser2015optimistic}
and \cite{maslov1992delinearization}.

\subsection{A "canonical" function call}
- Takes indeces (which are offset-corrected), number of elements per dimension.
- Show the universality of this with respect
\subsection{A new intrinsic}
\subsection{A new instruction}
\subsection{Modifying \gep}

\section{Caveats}
We do not take into account alignment, which is something that must be done
for this document to be complete.

\section{References: Links}

\begin{itemize}
    \item \href{https://chapel-lang.org/docs/1.13/_downloads/chapelLanguageSpec.pdf}{The Chapel language specification}.
    \item \href{http://www.j3-fortran.org/doc/year/04/04-007.pdf}{Fortran 2003 standard}.
\end{itemize}
\bibliography{references}
\bibliographystyle{plain}
\end{document}
